Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-06-11 intro.tex
[MGa] curvilinear abscissa -> curvilinear parameter

In some contexts, uncertainty arises from partial or imprecise knowledge of system parameters. This has led to a significant body of work on robust trajectory planning, particularly for unmanned aerial vehicles (UAVs), with the aim of minimizing sensitivity to state and input variations. For instance,~\cite{Brault:RobustTrajectoryPlanning:2021},~\cite{Giordano:TrajectoryGenerationMinimum:2018}, and~\cite{Brault:TubebasedTrajectoryOptimization:} propose tube-based and sensitivity-aware optimization frameworks that explicitly account for both input and closed-loop state sensitivity in the planning phase. Further developments integrate observability-aware planning into a unified multi-step optimization framework, as demonstrated in~\cite{Bohm:COPControlObservabilityaware:2022}.

>>>>

In some contexts, uncertainty arises from partial or imprecise knowledge of system parameters. This has led to a significant body of work on robust trajectory planning, particularly for unmanned aerial vehicles (UAVs), with the aim of minimizing sensitivity to state and input variations. For instance,~\cite{Brault:RobustTrajectoryPlanning:2021} and~\cite{Giordano:TrajectoryGenerationMinimum:2018} propose tube-based and sensitivity-aware optimization frameworks that explicitly account for both input and closed-loop state sensitivity in the planning phase. Further developments integrate observability-aware planning into a unified multi-step optimization framework, as demonstrated in~\cite{Bohm:COPControlObservabilityaware:2022}.

### 2025-06-11 performance_comparison.tex
[MGa] curvilinear abscissa -> curvilinear parameter

In this section we present some results obtained using the described robust planning frameworks. All planned trajectories refer to the sector of the Catalunya circuit shown in Figure~\ref{fig:track}.
The entire circuit is parametrized by the curve parameter $\alpha \in [0,1]$, and the selected sector corresponds to the interval $\left[0.70, 0.77\right]$. Checkpoints are indicated by labels and are uniformly spaced by $\De\al=0.01$. This sector includes two distinct corners: a low-speed turn from $\al = 0.72$ to $\al = 0.73$, and a high-speed turn from $\al = 0.75$ to $\al = 0.76$, allowing for the evaluation of vehicle behavior across different dynamic scenarios.

@@

Observing the left panel of Figure~\ref{fig:ol_sensitivities}, it is evident that the trajectories corresponding to configurations with lower values of $\ga^\textrm{TLC}$ yield smaller lateral margins with respect to the track boundaries. In contrast, the trajectory associated with $\ga^\textrm{TLC} = 3$\,--- represented by the light green line \,--- remains significantly farther from the edges.
This behavior is consistent with the increased conservativeness introduced by higher values of $\ga^\textrm{TLC}$, which amplify the back-off term $\be^\textrm{TLC}$ and thereby enforce a larger safety margin from the track limits.
Chasing a trajectory associated with a larger safety margin leads to higher sector times; spanning $\ga^\textrm{TLC}$ in the range $\left[0,3\right]$ results in sector times that vary from 12.135\,s to 13.162\,s.

@@

A comparison between the trajectories obtained using different values of $\bP_0$ is shown in the right panel of Figure~\ref{fig:ol_sensitivities}.
The black line corresponds to the baseline case with $\bP_0 = \bar{\bP}_0$, while the red and the blue lines represent the cases with $\bP_0 = \frac{1}{4}\bar{\bP}_0$ and $\bP_0 = 4\bar{\bP}_0$, respectively. As reasonably expected, using a larger initial covariance matrix results in an inflated covariance matrix that reflects increased uncertainty after $H$ steps, which in turn leads to a higher back-off term. This is associated to the blue line, which travels closer to the centerline with respect to the other two lines.

@@

\begin{figure}[t]
	\centering
	\includegraphics[scale = 1]{Fig/ol_saturation.pdf}
	\caption{Open-Loop. Axle saturation $S_j(\bx,\bu)$ for front axle (j=1, first panel) and rear axle (j=2, second panel) with respect to the curvilinear abscissa $\al$. Nominal configuration in dashed black, TLC in green, FLC in red, TLC+FLC in blue.
	%The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robustified configurations TLC, FLC, and TLC+FLC, respectively.
	}
	\label{fig:ol_saturation}
\end{figure}

@@

First we show results obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.
%All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.
%---Delta key signals
The comparison between the nominal case and the three robustified cases is illustrated in Figure~\ref{fig:ol_telemetries}, in terms of the deviation of selected signals from to their nominal counterparts.
The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass (CoM) from the track centerline $\De e$ (fourth panel).
Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal.
These indicators are shown at three distinct vertical levels --- high, medium, and low --- corresponding respectively to positive, near-zero, and negative values of the nominal signal.

@@

\begin{figure}[b!]
	\centering
	\includegraphics[scale = 1]{Fig/cl_saturation.pdf}
	\caption{Closed-Loop. Axle saturation $S_j(\bx,\bu)$ for front axle (j=1, first panel) and rear axle (j=2, second panel) with respect to the curvilinear abscissa $\al$. Nominal configuration in dashed black, TLC in green, FLC in red, TLC+FLC in blue.
		%The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robustified configurations TLC, FLC, and TLC+FLC, respectively.
	}
	\label{fig:cl_saturation}
\end{figure}

An analogous analysis has been conducted using the closed-loop approach described in Section~\ref{sec:closed_loop_planning}, setting $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$.
The same key signals displayed for the previous case are shown in Figure~\ref{fig:cl_telemetries}, while the axle saturation ratios $S_j$ are illustrated in Figure~\ref{fig:cl_saturation}.
From one side, the results appear very similar to those obtained with the open-loop method. While the closed-loop method offers a more rigorous framework by enabling seamless covariance propagation over the entire lap, the observed agreement indicates that open-loop planning remains a valid and effective strategy, as long as the prediction horizon is appropriately selected.

@@

According to the latter interpretation, Figure~\ref{fig:olcl_traj_bands} shows two highlighted bands---red for the open-loop case and cyan for the closed-loop case---corresponding to the area swept by the back-off-extended-vehicle. Ultimately, the TLC enforces precisely that these bands remain entirely within the actual track limits. The comparison between the two regions illustrates how the additional conservativeness of the open-loop approach stems from larger back-off terms, i.e., a wider band that must stay within the track boundaries. The closed-loop method, on the other hand, benefits from the LQR-based feedback, which allows it to maintain smaller back-off terms and thus a narrower band, requiring a reduced safety margin from the track limits.
\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_bands.pdf}
	\caption{Back-off envelopes for the open-loop (red) and closed-loop (cyan) robust planners. Each shaded band represents the area swept by the back-off-extended vehicle and must lie entirely within the actual track boundaries.
The open-loop approach yields a wider envelope due to worst-case growth assumptions, whereas the closed-loop design, aided by time-varying LQR controller, maintains a narrower margin.}
	\label{fig:olcl_traj_bands}
\end{figure}

@@

The left panel of Figure~\ref{fig:traj_strings} shows random evolutions (red lines) for an LQR-controlled vehicle stabilized to follow the nominal optimal trajectory (black line). No robust constraints are introduced, and the controller implements the feedback policy from Step 2 of Subsection~\ref{sec:LQR}. The right panel, instead, shows the random evolutions (red lines) for the robustified LQR-based controller obtained in the Step 3 of the closed-loop approach described in Subsection~\ref{sec:CLrobustified}. The corresponding robustified mean trajectory is shown in black.
The comparison between the panels shows that the robustified controller is significantly more effective at satisfying the TLC, resulting in safer and more reliable trajectory tracking.

@@

\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_strings.pdf}
	\caption{Simulations with random initial conditions and Gaussian noise realizations. In the left panel, the reference trajectory (black curve) is the non-robust one, tracked using an LQR controller designed around it as per Step 2 of Subsection~\ref{sec:LQR}. In the right panel, the reference (black curve) is a robust trajectory obtained using the closed-loop optimization method as per Step 3 of Subsection~\ref{sec:CLrobustified}, which is tracked using the corresponding optimized controller.}
	\label{fig:traj_strings}
\end{figure}

>>>>

In this section we present some results obtained using the described robust planning frameworks. All planned trajectories refer to the sector of the Catalunya circuit shown in Fig.~\ref{fig:track}.
The entire circuit is parameterized by the curvilinear parameter $\alpha \in [0,1]$, and the selected sector corresponds to the interval $\left[0.70, 0.77\right]$. Checkpoints are indicated by labels and are uniformly spaced by $\De\al=0.01$. This sector includes two distinct corners: a low-speed turn from $\al = 0.72$ to $\al = 0.73$, and a high-speed turn from $\al = 0.75$ to $\al = 0.76$, allowing for the evaluation of vehicle behavior across different dynamic scenarios.

@@

Observing the left panel of Fig.~\ref{fig:ol_sensitivities}, it is evident that the trajectories corresponding to configurations with lower values of $\ga^\textrm{TLC}$ yield smaller lateral margins with respect to the track boundaries. In contrast, the trajectory associated with $\ga^\textrm{TLC} = 3$\,--- represented by the light green line \,--- remains significantly farther from the edges.
This behavior is consistent with the increased conservativeness introduced by higher values of $\ga^\textrm{TLC}$, which amplify the back-off term $\be^\textrm{TLC}$ and thereby enforce a larger safety margin from the track limits.
Chasing a trajectory associated with a larger safety margin leads to higher sector times; spanning $\ga^\textrm{TLC}$ in the range $\left[0,3\right]$ results in sector times that vary from 12.135\,s to 13.162\,s.

@@

A comparison between the trajectories obtained using different values of $\bP_0$ is shown in the right panel of Fig.~\ref{fig:ol_sensitivities}.
The black line corresponds to the baseline case with $\bP_0 = \bar{\bP}_0$, while the red and the blue lines represent the cases with $\bP_0 = \frac{1}{4}\bar{\bP}_0$ and $\bP_0 = 4\bar{\bP}_0$, respectively. As reasonably expected, using a larger initial covariance matrix results in an inflated covariance matrix that reflects increased uncertainty after $H$ steps, which in turn leads to a higher back-off term. This is associated to the blue line, which travels closer to the centerline with respect to the other two lines.

@@

\begin{figure}[t]
	\centering
	\includegraphics[scale = 1]{Fig/ol_saturation.pdf}
	\caption{Open-Loop. Axle saturation $S_j(\bx,\bu)$ for front axle (j=1, first panel) and rear axle (j=2, second panel) with respect to the curvilinear parameter $\al$. Nominal configuration in dashed black, TLC in green, FLC in red, TLC+FLC in blue.
	%The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robustified configurations TLC, FLC, and TLC+FLC, respectively.
	}
	\label{fig:ol_saturation}
\end{figure}

@@

First we show results obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.
%All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.
%---Delta key signals
The comparison between the nominal case and the three robustified cases is illustrated in Fig.~\ref{fig:ol_telemetries}, in terms of the deviation of selected signals from to their nominal counterparts.
The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass (CoM) from the track centerline $\De e$ (fourth panel).
Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal.
These indicators are shown at three distinct vertical levels --- high, medium, and low --- corresponding respectively to positive, near-zero, and negative values of the nominal signal.

@@

\begin{figure}[b!]
	\centering
	\includegraphics[scale = 1]{Fig/cl_saturation.pdf}
	\caption{Closed-Loop. Axle saturation $S_j(\bx,\bu)$ for front axle (j=1, first panel) and rear axle (j=2, second panel) with respect to the curvilinear parameter $\al$. Nominal configuration in dashed black, TLC in green, FLC in red, TLC+FLC in blue.
		%The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robustified configurations TLC, FLC, and TLC+FLC, respectively.
	}
	\label{fig:cl_saturation}
\end{figure}

An analogous analysis has been conducted using the closed-loop approach described in Section~\ref{sec:closed_loop_planning}, setting $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$.
The same key signals displayed for the previous case are shown in Fig.~\ref{fig:cl_telemetries}, while the axle saturation ratios $S_j$ are illustrated in Fig.~\ref{fig:cl_saturation}.
From one side, the results appear very similar to those obtained with the open-loop method. While the closed-loop method offers a more rigorous framework by enabling seamless covariance propagation over the entire lap, the observed agreement indicates that open-loop planning remains a valid and effective strategy, as long as the prediction horizon is appropriately selected.

@@

According to the latter interpretation, Fig.~\ref{fig:olcl_traj_bands} shows two highlighted bands---red for the open-loop case and cyan for the closed-loop case---corresponding to the area swept by the back-off-extended-vehicle. Ultimately, the TLC enforces precisely that these bands remain entirely within the actual track limits. The comparison between the two regions illustrates how the additional conservativeness of the open-loop approach stems from larger back-off terms, i.e., a wider band that must stay within the track boundaries. The closed-loop method, on the other hand, benefits from the LQR-based feedback, which allows it to maintain smaller back-off terms and thus a narrower band, requiring a reduced safety margin from the track limits.
\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_bands.pdf}
	\caption{Back-off envelopes for the open-loop (red) and closed-loop (cyan) robust planners. Each shaded band represents the area swept by the back-off-extended vehicle and must lie entirely within the actual track boundaries.
The open-loop approach yields a wider envelope due to worst-case growth assumptions, whereas the closed-loop design, aided by time-varying LQR controller, maintains a narrower margin.}
	\label{fig:olcl_traj_bands}
\end{figure}

@@

The left panel of Fig.~\ref{fig:traj_strings} shows random evolutions (red lines) for an LQR-controlled vehicle stabilized to follow the nominal optimal trajectory (black line). No robust constraints are introduced, and the controller implements the feedback policy from Step 2 of Subsection~\ref{sec:LQR}. The right panel, instead, shows the random evolutions (red lines) for the robustified LQR-based controller obtained in the Step 3 of the closed-loop approach described in Subsection~\ref{sec:CLrobustified}. The corresponding robustified mean trajectory is shown in black.
The comparison between the panels shows that the robustified controller is significantly more effective at satisfying the TLC, resulting in safer and more reliable trajectory tracking.

@@

\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_strings.pdf}
	\caption{100 simulations (each panel) with random initial conditions and Gaussian noise realizations. In the left panel, the reference trajectory (black curve) is the non-robust one, tracked using an LQR controller designed around it as per Step 2 of Subsection~\ref{sec:LQR}. In the right panel, the reference (black curve) is a robust trajectory obtained using the closed-loop optimization method as per Step 3 of Subsection~\ref{sec:CLrobustified}, which is tracked using the corresponding optimized controller.}
	\label{fig:traj_strings}
\end{figure}
